\section{Proof of Lemma~\ref{lm:hytm}}
\label{app:lm}
Suppose, by contradiction that $T_1$ and $T_2$ 
contend on the same base object in $E_1\cdot E_2$.
Let $p_1$ (and resp. $p_2$) be the process executing transaction $T_1$ (and resp. $T_2$).

If in $E_1$, $T_1$ performs a nontrivial event on a base object on which they contend, let $e_1$ be the last
event in $E_1$ in which $T_1$ performs such an event to some base object $b$ and $e_2$, the first event
in $E_2$ that accesses $b$ (note that by assumption, $e_1$ is a cached access).
Otherwise, $T_i$ only performs trivial events in $E_1$ to base objects (some of which may be uncached) on which it contends with $T_{2}$ in $E_1\cdot E_2$:
let $e_2$ be the first event in $E_2$ in which $E_2$ performs a nontrivial event to some base object $b$
on which they contend and $e_1$, the last event of $E_1$ in $T_1$ that accesses $b$ (possibly uncached).

Let $E_1'$ (and resp. $E_2'$) be the longest prefix of $E_1$ (and resp. $E_2$) that does not include
$e_1$ (and resp. $e_2$).
Since before accessing $b$, the execution is step contention-free for $T_1$, $E \cdot
E_1'\cdot E_2'$ is an execution of $\mathcal{M}$.
By assumption of lemma, $T_1$ and $T_2$ do not conflict in $E_1'\cdot E_2'$.
By construction, $E_1 \cdot E_2'$ is indistinguishable to $T_2$ from $E_1' \cdot E_2'$.
Hence, $T_1$ and $T_{2}$ are poised to apply contending events $e_1$ and $e_2$ on $b$ in the execution
$\tilde E=E_1' \cdot E_2'$.

We now consider two cases:
\begin{enumerate}
\item 
($e_1$ is a nontrivial event)
After $\tilde E\cdot e_1$, $b$ is contained in the tracking set of process
$p_1$ in exclusive mode and in the extension $\tilde E\cdot e_1 \cdot e_2$, we have that
$\tau_1$ is invalidated. Thus, by Remark~\ref{re:traborts}, transaction $T_1$ must return $A_1$ 
in any extension of $E\cdot e_1\cdot e_2$---a contradiction
to the assumption that $\mathcal{M}$ is progressive.   
\item
($e_1$ is a trivial event)
Recall that $e_1$ may be potentially uncached.
Consider the execution $\tilde E\cdot e_2$ following which $b$ is contained in the tracking set of process
$p_{2}$ in exclusive mode. Clearly, we have an extension $\tilde E\cdot e_2 \cdot e_1$ in which
$\tau_{2}$ is invalidated. Thus transaction $T_{2}$ must return $A_{2}$ in any extension of $E\cdot e_2\cdot e_1$---a contradiction
to the assumption that $\mathcal{M}$ is progressive.   
\end{enumerate}
%
