\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\journal{Journal of \LaTeX\ Templates}
\usepackage{multicol}
%\usepackage{amsmath}
\usepackage{framed}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{float}
\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{tikz}
\usepackage{macros}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{ioa_code}

\newcommand{\comnospace}{$\triangleright$}
\newcommand{\com}{\comnospace\ }
\usepackage{tabto}
\newcommand{\medcom}[1]{\tabto{5cm} \com \mbox{#1}}

\usepackage{listings}
\lstset{numbers=left,numberblanklines=false}

\lstdefinestyle{customc}{
	belowcaptionskip=1\baselineskip,
	breaklines=true,
	frame=L,	
	xleftmargin=\parindent,
	language=C,
	showstringspaces=false,
	escapeinside={//}{\^^M},
	%    basicstyle=\LSTfont,
	basicstyle=\scriptsize\ttfamily,
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{gray!60!black},
	identifierstyle=\color{blue!50!black},
	stringstyle=\color{orange},
	numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
	numbersep=4pt,                   % how far the line-numbers are from the code
	numberstyle=\tiny\color{black},  % the style that is used for the line-numbers
	otherkeywords={then,word,process_local,type,xbegin,xabort,xend}
}

\def\P{\ensuremath{\mathcal{P}}}
\def\DP{\ensuremath{\Diamond\mathcal{P}}}
\def\DS{\ensuremath{\Diamond\mathcal{S}}}
%\def\T{\ensuremath{\mathcal{T}}}
\def\Time{\mathbb{T}}
\def\S{\ensuremath{\mathcal{S}}}
\def\D{\ensuremath{\mathcal{D}}}
\def\W{\ensuremath{\mathcal{W}}}
\def\A{\ensuremath{\mathcal{A}}}
\def\B{\ensuremath{\mathcal{B}}}
\def\F{\ensuremath{\mathcal{F}}}
\def\R{\ensuremath{\mathcal{R}}}
\def\N{\ensuremath{\mathcal{N}}}
\def\I{\ensuremath{\mathcal{I}}}
\def\O{\ensuremath{\mathcal{O}}}
\def\Q{\ensuremath{\mathcal{Q}}}
\def\K{\ensuremath{\mathcal{K}}}
\def\L{\ensuremath{\mathcal{L}}}
\def\M{\ensuremath{\mathcal{M}}}
\def\V{\ensuremath{\mathcal{V}}}
\def\E{\ensuremath{\mathcal{E}}}
\def\C{\ensuremath{\mathcal{C}}}
\def\T{\ensuremath{\mathcal{T}}}
\def\X{\ensuremath{\mathcal{X}}}
\def\Y{\ensuremath{\mathcal{Y}}}
\def\Nat{\ensuremath{\mathbb{N}}}
\def\Om{\ensuremath{\Omega}}
\def\ve{\varepsilon}
\def\fd{failure detector}
\def\cfd{\ensuremath{?\P+\DS}}
\def\afd{timeless}
\def\env{\ensuremath{\mathcal{E}}}
%\def\faulty{unreliable}
\def\bounded{one-shot}
\def\cons{\textit{cons}}
\def\val{\textit{val}}
\def\code{\textit{code}}

\def\HSS{\mathit{h}}
\def\argmin{\mathit{argmin}}
\def\proper{\mathit{proper}}
\def\content{\mathit{content}}
\def\Level{\mathit{L}}
\def\Blocked{\mathit{Blocked}}
\def\Set{\mathit{Set}}
\def\TS{\mathit{TS}}
\def\shared{\mathit{shared}}
\def\exclusive{\mathit{exclusive}}

\newcommand{\correct}{\mathit{correct}}
\newcommand{\RNum}[1]{\uppercase\expandafter{\romannumeral #1\relax}}
\newcommand	{\faulty}{\mathit{faulty}}
\newcommand{\infi}{\mathit{inf}}
\newcommand{\live}{\mathit{live}}
\newcommand{\true}{\mathit{true}}
\newcommand{\false}{\mathit{false}}
\newcommand{\stable}{\mathit{Stable}}
\newcommand{\setcon}{\mathit{setcon}}
\newcommand{\remove}[1]{}

\newcommand{\Wset}{\textit{Wset}}
\newcommand{\Rset}{\textit{Rset}}
\newcommand{\Dset}{\textit{Dset}}

%\newcommand{\parts}{\textit{parts}}
\newcommand{\txns}{\textit{txns}}

\newcommand{\Read}{\textit{read}}
\newcommand{\Write}{\textit{write}}
\newcommand{\TryC}{\textit{tryC}}
\newcommand{\TryA}{\textit{tryA}}
\newcommand{\ok}{\textit{ok}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newenvironment{proofsketch}[1][Proof sketch]{\noindent\textbf{#1.} }{\hfill $\Box$\\[2mm]}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{claim}[theorem]{Claim}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{Cost of Concurrency in Hybrid Transactional Memory}

\author{Elsevier\fnref{myfootnote}}
\address{Radarweg 29, Amsterdam}
\fntext[myfootnote]{Since 1880.}

%% or include affiliations in footnotes:
\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
\ead[url]{www.elsevier.com}

\author[mysecondaryaddress]{Global Customer Service\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{support@elsevier.com}

\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
State-of-the-art \emph{software transactional memory (STM)} implementations achieve 
good performance by carefully avoiding the overhead of \emph{incremental validation}
(i.e., re-reading previously read data items to avoid inconsistency) while
still providing \emph{progressiveness} (allowing transactional aborts only due to \emph{data conflicts}).
Hardware transactional memory (HTM) implementations promise even better performance, 
but offer no progress guarantees.
Thus, they must be combined with STMs, leading to \emph{hybrid} TMs (HyTMs)
in which hardware transactions must be \emph{instrumented} (i.e., access metadata) 
to detect contention with software transactions.

We show that, unlike in progressive STMs, software transactions in progressive HyTMs
cannot avoid incremental validation.
In fact, this result holds even if hardware transactions can \emph{read} metadata 
\emph{non-speculatively}. 
We then present \emph{opaque} HyTM algorithms providing \emph{progressiveness for a subset of transactions} 
that are  optimal in terms of hardware instrumentation. 
We explore the concurrency vs. hardware instrumentation vs. software validation
trade-offs for these algorithms.
Our experiments with Intel and IBM POWER8 HTMs   
seem to suggest that (i) the \emph{cost of concurrency} also exists in practice, 
(ii) it is important to implement HyTMs that provide progressiveness for a maximal set of transactions without incurring high hardware instrumentation overhead or
using global contending bottlenecks and (iii) 
there is no easy way to derive more efficient HyTMs by taking advantage of non-speculative accesses within hardware.

\end{abstract}

\begin{keyword}
\texttt{elsarticle.cls}\sep \LaTeX\sep Elsevier \sep template
\MSC[2010] 00-01\sep  99-00
\end{keyword}

\end{frontmatter}

\linenumbers

\input{intro}
\input{model_alt}
%\input{spec}
%\input{algos}
%
%\input{Evaluation}
%
%\input{related}

\section*{Acknowledgements}
Computations were performed on the SOSCIP Consortium’s Agile computing platform. 
SOSCIP is funded by the Federal Economic Development Agency of Southern Ontario, the Province of Ontario, IBM Canada Ltd., 
Ontario Centres of Excellence, Mitacs and 15 Ontario academic member institutions. 
Trevor Brown received funding for this work from the Natural Sciences and Engineering Research Council of Canada.
This work was performed while Trevor Brown was a student at the University of Toronto.


\section*{References}

\bibliography{refs}

\end{document}