%!TEX root = htm.tex
\section{Related work and discussion}
\label{sec:rel}
%
\subsection{HyTM implementations and complexity}
Early HyTMs like the ones described in \cite{damronhytm, kumarhytm} provided progressiveness, but
subsequent HyTM proposals like PhTM~\cite{phasedtm} and HybridNOrec~\cite{hybridnorec} sacrificed progressiveness for lesser instrumentation overheads.
However, the clear trade-off in terms of concurrency vs. instrumentation for these HyTMs have not been studied in the context of currently available HTM
architectures. This instrumentation cost on the fast-path was precisely characterized in \cite{hytm14disc}.
In this paper, we proved the inherent cost of concurrency on the slow-path thus establishing a surprising, 
but intuitive complexity separation between progressive STMs and HyTMs.
Moreover, to the best of our knowledge, this is the first work to consider the theoretical foundations of the cost of concurrency in 
HyTMs in theory and practice (on currently available HTM architectures).
Proof of Theorem~\ref{th:impossibility} is based on the analogous proof for step complexity of STMs that are \emph{disjoint-access parallel}~\cite{prog15-pact, tm-book}.
Our implementation of Hybrid NOrec follows \cite{hynorecriegel}, which additionally proposed the use of direct accesses
in fast-path transactions to reduce instrumentation overhead in the AMD Advanced Synchronization Facility (ASF) architecture.

\subsection{Beyond the two path HyTM approach}
\vspace{1mm}\noindent\textit{Employing an uninstrumented fast fast-path.}
We now describe how every transaction may first be executed in a ``fast'' fast-path with almost no instrumentation
and if unsuccessful, may be re-attempted in the fast-path and subsequently in slow-path.
Specifically, we transform an opaque HyTM $\mathcal{M}$ to an opaque
HyTM $\mathcal{M}'$ using a shared \emph{fetch-and-add} metadata base object $F$ that slow-path updating transactions
increment (and resp. decrement) at the start (and resp. end). In $\mathcal{M}'$, a ``fast'' fast-path transaction checks first if $F$ is $0$
and if not, aborts the transaction; otherwise the transaction is continued as an uninstrumented hardware transaction.
The code for the fast-path and the slow-path is identical to $\mathcal{M}$.
%Assuming the hardware transactions do not perform any direct accesses, opacity is immediate~\cite{brownfaster15}.

\subsection{Other approaches}
Recent work has investigated fallback to \emph{reduced} hardware transactions~\cite{MS13}
in which an all-software slow-path is augmented using a slightly faster slow-path 
that is optimistically used to avoid running some transactions on the true software-only slow-path.
Amalgamated lock elision (ALE) was proposed in \cite{ale15} which improves over TLE
by executing the slow-path as a series of segments, each of which is a dynamic length hardware transaction.
Invyswell~\cite{Calciu14} is a HyTM design with multiple hardware and software modes of execution that gives flexibility to avoid instrumentation overhead in uncontended executions.
We remark that such multi-path approaches may be easily applied to each of the Algorithms proposed in this paper. However, 
in the search for an efficient HyTM, it is important to strike the fine balance between concurrency, hardware instrumentation and software validation cost.
Our lower bound, experimental methodology and evaluation of HyTMs provides the first clear characterization of these trade-offs in both Intel and POWER8 architectures. 
%